---
title: 'Reproducible Research: Peer Assessment 1'
author: "Markus Friede Hens"
date: "Monday, November 03, 2014"
output: html_document
---

##Assumptions
The script assumes that the activity data is present in the current working directory under the name `activity.csv`. 
The file must contain the folowing features:

Feature|Type|Description
-------|----|-----------
steps|int|No. of steps taken
date|Date (YYYY-MM-DD)|Date of day
interval|int|Interval in day given by date, from 0:00 to 23:55 in 5 minute steps

###Loading and Preprocessing the data
```{r}
activitydata <- read.csv("activity.csv", stringsAsFactors = FALSE)
activitydata$date <- as.Date(activitydata$date)
```

###What is the mean total number of steps taken per day?

Aggregate the number of steps per day, based on this mean and median can be easily calculated. `NA`s are ignored.

```{r}
sumPerDate <- aggregate(activitydata$steps, list(Date = activitydata$date), FUN=sum, na.rm=TRUE)
meanNoOfStepsPerDay <- mean(sumPerDate$x, na.rm=TRUE)
medianNoOfStepsPerDay <- median(sumPerDate$x, na.rm=TRUE)
```

```{r echo=FALSE}
hist(sumPerDate$x, xlab="Steps per day", main = "Histogram of steps per day")
```

The sample mean is `r format(round(meanNoOfStepsPerDay, 2), nsmall=2)` while the sample median is `r medianNoOfStepsPerDay`.

### What is the average daily activity pattern?
Calculate the mean per interval. `NA`s are ignored.

```{r}
meanPerInterval = aggregate(activitydata$steps, list(interval = activitydata$interval), mean, na.rm=TRUE)
```

```{r echo=FALSE}
plot(meanPerInterval$interval, meanPerInterval$x, type="l", main ="Avg. of steps per 5-minute interval", xlab="Interval", ylab="Avg. no of. steps")
```

###Imputing missing values
The total number of rows in the data set containing `NA`s is `r sum(is.na(activitydata$steps))`

`NA`s are replaced by the average number of steps for that interval over all days.

``` {r}
activityDataNoNA <- activitydata
rowsWithNA <- which(is.na(activitydata$steps))
activityDataNoNA[rowsWithNA, "steps"] <- round(merge(activitydata[rowsWithNA,], meanPerInterval)$x)
```

Aggregate the number of steps per day, again, this time on the cleaned data set containing no `NA`s, and calculate mean and median again.

```{r}
sumPerDateNoNA <- aggregate(activityDataNoNA$steps, list(Date = activityDataNoNA$date), FUN=sum)
meanNoOfStepsPerDayNoNA <- mean(sumPerDateNoNA$x)
medianNoOfStepsPerDayNoNA <- median(sumPerDateNoNA$x)
```

```{r echo=FALSE}
hist(sumPerDateNoNA$x, xlab="Steps per day", main = "Histogram of steps per day (imputed)")
```

The sample mean of the cleaned data set is `r format(round(meanNoOfStepsPerDayNoNA, 2), nsmall=2)` while the sample median is `r format(medianNoOfStepsPerDayNoNA)`. The mean increased by `r format(round(meanNoOfStepsPerDayNoNA-meanNoOfStepsPerDay, 2), nsmall=2)`, while the median increased by `r format(medianNoOfStepsPerDayNoNA-medianNoOfStepsPerDay)`.

###Are there differences in activity patterns between weekdays and weekends?
Add information wether a given date is a weekday or not to the data set.
Calculate the mean per interval and weekday/weekend on the augmented data set.

``` {r} 
weekday <- as.factor(vapply(weekdays(activityDataNoNA$date), FUN=function(x) ifelse(x %in% c("Samstag", "Sonntag"), "weekend", "weekday"), FUN.VALUE="w"))

activityDataNoNA <- cbind(activityDataNoNA, weekday)

meanPerIntervalNoNA = aggregate(activityDataNoNA$steps, list(interval = activityDataNoNA$interval, weekday=activityDataNoNA$weekday), mean)
```

```{r echo=FALSE}
par(mfrow=c(1,2))
plot(meanPerIntervalNoNA[meanPerIntervalNoNA$weekday=="weekday",]$interval, meanPerIntervalNoNA[meanPerIntervalNoNA$weekday=="weekday",]$x, type="l", main ="Avg. of steps\nper 5-minute interval\nWeekday", xlab="Interval", ylab="Avg. no of. steps")
plot(meanPerIntervalNoNA[meanPerIntervalNoNA$weekday=="weekend",]$interval, meanPerIntervalNoNA[meanPerIntervalNoNA$weekday=="weekend",]$x, type="l", main ="Avg. of steps\nper 5-minute interval\nWeekend", xlab="Interval", ylab="Avg. no of. steps")
```

The plots show a clear difference in activity patterns between weekdays and weekends. On weekdays, there is a spike at around interval 7:50, lesser spikes around lunch time, and again in the afternoon and evening, with limited activity in between. In the weekend, activity levels are generally higher, and more spread out over the day. A possible explanation could be  person with an office job, that goes to work, goes out for lunch, home again, to do something with his famlity. On the weekends, the jobroutine is replaced by freetime activities, obviously involving more movement, maybe a runner.
